---
title: "Coursera - Activity Monitoring - Reproducible Research - Assignment -1"
author: "Rajathithan Rajasekar"
date: "March 17, 2017"
output: html_document
keep_md: true
---

**Declaring the Libraries that are required**

```{r}
library(lattice)
```

**Read the activity.csv file if it is present in the local directory.**

```{r}
if(!exists("activity.csv")){
        activitylist <- read.csv('activity.csv',na.strings = c("NA"))
        
}
```

**Basic Analysis to check the data**

```{r}
summary(activitylist)
head(activitylist)
tail(activitylist)
str(activitylist)
```

**There are 3 columns - Steps, date and interval. The steps column contains "NAs" , Data is taken over of 2 month period from October 2012 to November 2012 (61 days ) and the interval timeframe is from morning till night for every 5 mins.** 

**Transforming the data into a format suitable for our Analysis**

**Date values are converted from factor format to date format and interval is converted from int format to factor format.** 

```{r}
activitylist$date = as.Date(activitylist$date, format = "%Y-%m-%d")
activitylist$interval = as.factor(activitylist$interval)
str(activitylist)
```

**Creating a Data by Omitting the 'NA' rows**

```{r}
alistNA <- na.omit(activitylist)
head(alistNA)
```


**To find total number of steps taken per day **
```{r}
stepsperday<- tapply(alistNA$steps, alistNA$date, sum)
stepsperday
```

**Histogram on total number of steps taken per day**

```{r}
hist(stepsperday, col="orange",
     xlab="Number of Steps",
     ylab = "Frequency",
     main="Total Number of Steps Taken Each Day")
```


**Mean of the total steps taken per day**
```{r}
meanstepsperday <- mean(stepsperday)
meanstepsperday
```

**Median of the total steps taken per day**
```{r}
medianstepsperday <- median(stepsperday)
medianstepsperday
```


**To find Average no. steps taken across all days **
```{r}
avgstepsperday<- tapply(alistNA$steps, alistNA$interval, mean)
avgstepsperday
names(avgstepsperday)
```

**A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r}
plot(names(avgstepsperday), avgstepsperday, type="l",col="red", xlab="5-min interval", ylab="Average no. of steps across all days",main="Time Series Plot")
```

**The 5-minute interval which contains the maximum number of steps  on an average across all the days in the dataset**

```{r}
maxavg<- max(avgstepsperday)
maxavg
maxinterval<- names(avgstepsperday)[which(avgstepsperday==max(avgstepsperday))]
maxinterval
```

**Total no of NAs in the dataset **
```{r}
totalna<- sum(is.na(activitylist$steps))
totalna
```

**Modify the data by imputing missing NA values from avgsteps for all days during that time interval**

```{r}
Modifieddata <- activitylist

for(i in which(is.na(Modifieddata$steps))) {
    #Get the NA index 
    int <- Modifieddata$interval[i]
    #get the mean value for that 5-minute interval
    avgStepsForInt <- avgstepsperday[which (names(avgstepsperday) == int )]
    #replace the NA with the mean value
    Modifieddata$steps[i] <- avgStepsForInt
}
```

**Total no of NAs in the modified dataset, // ideally this should be Zero :) //**
```{r}
totalna<- sum(is.na(Modifieddata$steps))
totalna
```

**To find total number of steps taken per day with the modified data set**
```{r}
stepsperday<- tapply(Modifieddata$steps, Modifieddata$date, sum)
stepsperday
```

**Histogram on total number of steps taken per day with the modified data set**

```{r}
hist(stepsperday, col="orange",
     xlab="Number of Steps",
     ylab = "Frequency",
     main="Total Number of Steps Taken Each Day")
```

**Mean of the total steps taken per day on the modified data set**
```{r}
meanstepsperday <- mean(stepsperday)
meanstepsperday
```

**Median of the total steps taken per day on the modified data set**
```{r}
medianstepsperday <- median(stepsperday)
medianstepsperday
```

**OBSERVATION: The median and mean got closer**

**Creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day**

```{r}
Modifieddata <- transform(Modifieddata, dayType = ifelse(weekdays(Modifieddata$date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
#make it a factor
Modifieddata <- transform(Modifieddata, dayType = factor(dayType))
head(Modifieddata)
```

**Creating a panel plot containing a time series plot with type = "l" of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days and weekend days (y-axis).**

```{r}
plotData1 <- aggregate(steps ~ interval + dayType, data=Modifieddata, mean)
head(plotData1)
xyplot(steps ~ interval|dayType, data = plotData1,
       type = "l", layout = c(1,2),col="red",
       xlab="5-Minute Interval", ylab = "Average Number of Steps Taken",
       main="Average Number of Steps by 5-Minute Intervals (Weekdays vs. Weekends)")
```
